Continuous Deductive Reasoning over Streamed RDF Data

### 2015-07-14 content/contribution/continuous-datalog.tex
moved the proof of the theorem from appendix to section on c-datalog, then commented out to be re-introduced if desired.  Reformatted appendix to a set of figures, being general case, base case and inductive case
Revised defn of C-Datalog, added initial defn of WC-Datalog

C-Datalog differs from conventional Datalog through the addition of
sources of continuously generated or reported facts (i.e streams), and
the application of sliding window semantics over those sources,
thereby providing for the interpretation of Continuous Datalog
programs as sequences of instantaneous states that are each valid
positive Datalog programs.  As such, these states each inherit the
semantics of positive Datalog.  Furthermore, it involves the
continuous derivation of entailments from this sequence of
instantaneous states as a set of streams of facts; these streams share
a data model with the streams of continuous queries, as defined by the
stream processing community, thereby allowing the integration of
Continuous Datalog programs into stream processing pipelines.

In addition, we show that these streams may be manipulated to retrieve
the instantaneous sets of entailments produced by the more
semantically well-defined offerings from the reasoning community.

%%% what does this mean? ^

@@

A Time-Annotated Datalog program (henceforth abbreviated TA-Datalog)
$\Pi^T = \langle R, F^T\rangle$ is a tuple which comprises a finite
set of rules $R$ and a finite set of fact instances $F^T \subseteq
\hat{H} \times \mathbb{T}$. Fact instances are as defined in
\refdef{continuous datalog: fact instance}, and rules are of the same
form as in conventional Datalog programs.

When considering a TA-Datalog program, in which there may be many
distinct instances of a given fact, temporal information regarding
when instances of that fact became true would be lost if such a
program were to entail sets of plain facts without temporal
annotations. We therefore extend the model theory for Datalog from
\refdef{continuous datalog: datalog program} by considering the
Herbrand base of all fact instances $\hat{H}^T$ (all facts at all
instants). As before, a fact instance $e$ is entailed by a program
$\Pi^T$ iff each interpretation satisfying every clause in $\Pi^T$
also satisfies $e$, written $\Pi^T \models e$. The {\em entailment of the
program} $\Pi^T$ is therefore:
\[ cons^T(\Pi^T) = \{ e \in \hat{H}^T \mid \Pi^T \models e \} \]

@@

Reconsidering the example TA-Datalog program above, we may convert it
into the following equivalent conventional Datalog program $\Pi$:
\begin{align*}
\Pi &= \langle \{ c \Leftarrow a \land b \}, strip(\{ \langle a, 1 \rangle, \langle a, 2 \rangle, \langle b, 1 \rangle, \langle b, 2 \rangle \}) \rangle \\
    &= \langle \{ c \Leftarrow a \land b \}, \{ a, b \} \rangle
\end{align*}
Trivially, the entailment of this program $\Pi$ will therefore be:
\begin{align*}
cons(\Pi) &= strip(\{\langle c, 1 \rangle, \langle c, 2 \rangle\}) \\
 &= \{ c \}  
\end{align*}
\end{definition}

@@

A {\em stream} is an unbounded sequence of fact instances ordered by
their instants; the instants are non-decreasing in the stream, such
that for any two fact instances $\langle f_i, t_i \rangle$ and
$\langle f_j, t_j\rangle$ where the former appears on the stream
  before the latter, $t_i \leqslant t_j$.
\end{definition}

@@

A window of range ${x \in \mathbb{N}}$ over a stream $S$ at an instant
${i \in \mathbb{T}}$ is the \emph{set} of fact instances that occur in
$S$ during the interval ${(i-x,i]}$, written ${W^{S,x}_{i}}$, and
defined as follows:
\[ W^{S,x}_i = \{ \langle f, t \rangle \mid \langle f, t \rangle \in S \land (i-x) < t \leqslant i \}\]  
\end{definition}

@@

A Continuous Datalog program (henceforth abbreviated C-Datalog)
$\Pi^C = \langle R, F^0, S, r \rangle$ is a tuple which comprises a
finite set of rules $R$, a finite set of static fact instances $F^0$
(all of which have the instant $0$), a stream $S$ and a range
${r \in \mathbb{N}}$ for a window over $S$.

%This specifies that the streamed instances from $S$
%are to be considered valid from their timestamp for the duration
%represented by $r$.

We consider the entailment of the program $\Pi^C$ not just
as a set of fact instances (as for TA-Datalog), but in three parts: a
set of fact instances $E^0$ entailed by the fact instances in $F^0$, a
stream $P$ of fact instances entailed by the fact instances in $F^0$
and S, and a stream $N$ of fact instance negations: $\langle R, F^0,
S, r \rangle \models \langle E^0, P, N \rangle$. $P$ and $N$ can be
considered to be the continuous reasoning equivalents of the streams
produced by the \emph{IStream} and \emph{DStream} operators
respectively of the \emph{CQL} continuous query language \citep{CQL},
which create streams based on the additions to and deletions from a
relation.

@@

Given a C-Datalog program $\Pi^C$, we can

The truth value of a $\Pi^C$ is instantaneous, with the the truth
function of Continuous Datalog $T^{CD}$ taking both a program ${\Pi^C}$
and an instant ${i \in \mathbb{N}}$ as arguments.

The state of a $\Pi^C$ at a given instant $i$ is the time-annotated
Datalog program ${\Pi^C_i = \langle R, F^0 \cup W^{S,r}_{i} \rangle}$,
where ${F^0 = \{ f | 0 : f \in F \}}$ is the set of persistent fact instances produced by
annotating each fact in $f$ in $F$ as becoming true at the first recognised instant.

As both $F$ and ${W_i = \{ strip(x) : x \in W^{S,r}_{i} \}}$ are sets of conventional
Datalog facts, the tuple ${\langle R, F \cup W_i}$
constitutes a valid Datalog program that may be evaluated by the truth
function of Datalog $T^{D}$, the result of which is the definition for the
instantaneous truth of $\Pi^C$ at instant $i$ by $T^{CD}$.
\begin{multline*}
T^{CD} \left( \langle R, F, S, r \rangle, i \right) = \\
	T^{D} \left( \langle R, F \cup \{ strip(x) : x \in W^{S,r}_{i} \} \right)
\end{multline*}

>>>>

C-Datalog differs from conventional Datalog through the addition of
sources of continuously generated or reported facts (i.e streams), and
the application of sliding window semantics over those sources,
thereby providing for the interpretation of C-Datalog programs as
sequences of instantaneous states that are each valid positive Datalog
programs.  As such, these states each inherit the semantics of
positive Datalog.  Furthermore, it involves the continuous derivation
of entailments from this sequence of instantaneous states as a stream
of facts; these streams share a data model with the streams of
continuous queries, as defined by the stream processing community,
thereby allowing the integration of C-Datalog programs into stream
processing pipelines. Finally, we show that these streams may be
manipulated to retrieve the instantaneous sets of entailments produced
by the equivalent Datalog programs.

@@

A {\em Time-Annotated Datalog program} (henceforth abbreviated
TA-Datalog program) $\Pi^T = \langle R, F^T\rangle$ is a tuple which
comprises a finite set of rules $R$ and a finite set of fact instances
$F^T \subseteq \hat{H} \times \mathbb{T}$. Fact instances are as
defined in \refdef{continuous datalog: fact instance}, and rules are
of the same form as in conventional Datalog programs.

When considering a TA-Datalog program, in which there may be many
distinct instances of a given fact, temporal information regarding
when instances of that fact became true would be lost if such a
program were to entail sets of plain facts without temporal
annotations. We therefore extend the model theory for Datalog from
\refdef{continuous datalog: datalog program} by considering the
Herbrand base of all fact instances $\hat{H}^T$ (all facts with all
instants). As before, a fact instance $e$ is entailed by a program
$\Pi^T$ iff each interpretation satisfying every clause in $\Pi^T$
also satisfies $e$, written $\Pi^T \models e$. The {\em entailment of the
TA-Datalog program} $\Pi^T$ is therefore:
\[ cons^T(\Pi^T) = \{ e \in \hat{H}^T \mid \Pi^T \models e \} \]

@@

Reconsidering the example TA-Datalog program above, we may reduce it
to the following equivalent conventional Datalog program $\Pi$:
\begin{align*}
\Pi &= \langle \{ c \Leftarrow a \land b \}, strip(\{ \langle a, 1 \rangle, \langle a, 2 \rangle, \langle b, 1 \rangle, \langle b, 2 \rangle \}) \rangle \\
    &= \langle \{ c \Leftarrow a \land b \}, \{ a, b \} \rangle
\end{align*}
Trivially, the entailment of this program $\Pi$ will therefore be:
\begin{align*}
cons(\Pi) &= strip(\{\langle c, 1 \rangle, \langle c, 2 \rangle\}) \\
 &= \{ c \}  
\end{align*}
\end{definition}

@@

A {\em stream} $S$ is an unbounded, ordered sequence of fact
instances. We indicate that a fact instance $\langle f, t\rangle$
appears in $S$ by writing $\langle f, r \rangle \in S$. We indicate
that (in a given stream) fact instance $f_i$ appears before fact
instance $f_j$ by writing $f_i \prec f_j$.

Streams may be (but are not necessarily) ordered by the instants in
their fact instances, with the instants non-decreasing in the stream,
such that for any two fact instances $\langle f_i, t_i \rangle$ and
$\langle f_j, t_j\rangle$ where $\langle f_i, t_i \rangle \prec
\langle f_j, t_j \rangle$, $t_i \leqslant t_j$; we refer to such
streams as {\em in-order streams}, and to streams whose fact instances
are not ordered by their instants as {\em out-of-order streams}.
\end{definition}

@@

A window of range ${x \in \mathbb{N}}$ over a stream $S$ at an instant
${i \in \mathbb{T}}$ is the \emph{set} of fact instances that occur in
$S$ with instances in the interval ${(i-x,i]}$, written
  ${W^{S,x}_{i}}$, and defined as follows:
\[ W^{S,x}_i = \{ \langle f, t \rangle \mid \langle f, t \rangle \in S \land (i-x) < t \leqslant i \}\]  
\end{definition}

@@

We initially take a {\em Continuous Datalog program} (henceforth a
C-Datalog program) $\Pi^C = \langle R, F^0, S \rangle$ to be a tuple
which comprises a finite set of rules $R$, a finite set of static fact
instances $F^0$ (all of which have the instant $0$) and an in-order
stream of fact instances $S$.

The model theory for a such a C-Datalog program broadly follows that
for a TA-Datalog program; we again take the Herbrand base of all fact
instances $\hat{H}$. Disregarding the order of fact instances in $S$,
a fact instance $e$ is entailed by a program $\Pi^C$ iff each
interpretation satisfying every clause in $\Pi^C$. However, we take
the entailment of the program $\Pi^C$ to be an in-order stream $E^S$,
where the order in which entailed fact instances appear in $E^S$
depends on the order in which the fact instances in their
justifications appear in $S$. Given that that instant of an entailed
fact instance is the maximum of the instants in the fact instances
that form the justification for that entailed fact instance (as in
\refdef{continuous datalog: instance justification}), $E^S$ will also
be an in-order stream.

However, this definition of a C-Datalog program does not take into
account any notion of windowing over the stream $S$. We define a {\em
  Windowed Continuous Datalog program} $\Pi^{WC} = \langle R, F^0, S,
r \rangle$ to be a tuple with components as defined above for a
C-Datalog program, but with the addition of $r \in \mathbb{N}$ for the
range of the window over $S$, which effectively specifies that the
fact instances from $S$ are to contribute to the entailments of $\Pi^{WC}$
from their instant for a duration of $r$.

We consider the entailment of the WC-Datalog program $\Pi^{WC}$ not as
a single stream of fact instances (as for C-Datalog), but in three
parts: a set of fact instances $E^0 = cons^T(\langle R, F^0\rangle$
that is the entailment of the TA-Datalog program $\langle R,
F^0\rangle$, a stream $P$ of fact instances entailed by the fact
instances in $F^0$ and $S$, and a stream $N$ of fact instance negations:
$\langle R, F^0, S, r \rangle \models \langle E^0, P, N \rangle$. $P$
and $N$ can be considered to be the continuous reasoning equivalents
of the streams produced by the \emph{IStream} and \emph{DStream}
operators respectively of the \emph{CQL} continuous query language
\citep{CQL}, which create streams based on the additions to and
deletions from a relation.

@@

We can reduce a C-Datalog program $\Pi^C = \langle R, F^0, S, r
\rangle$ to its TA-Datalog equivalent by taking the window $W^{S,r}_i$
over stream $S$ of range $r \in \mathbb{N}$ at instant $i \in
\mathbb{T}$; we refer to this as the {\em instantaneous state} of the
C-Datalog program, written as $\Pi^C_i = \langle R, F^0 \cup W^{s,r}_i
\rangle$. As previously shown in \refdef{continuous datalog: ta
  datalog program}, this may then be further reduced to the Datalog
program $\Pi = \langle R, strip(F^0 \cup W^{S,r}_i) \rangle$.
